---
title: "Machine Learning, Text and Networks Demo"
output: github_document
---

### Word Clouds

We've got a dataset of a Gallup survey from 2014 that asks an open ended question about how people would feel if Hilary Clinton were to become president.  What is the most positive thing and what is the worst thing about it?  First, we need to do some basic read in and data manipulation.  It's in a weird format (an SPSS portable file) so I had to do some cumbersome gymnastics to get it into shape.  Not elegant, but very real.  

We also create a new roll-up variable of ideology, in which we take the 7 point ideology scale they asked and collapse it into 3 categories: Conservative, Moderate and Liberal.

```{r}
library(haven) # for reading SPSS data files
library(tidyverse)

d1 <- read_por("g2014tr0315.por")

d2 <- d1 %>% 
   select("ID", "WTFCTR", "LANGUAGE", "P862_1_T", "P862_2_T", "P862_3_T", "P863_1_T", "P863_2_T", "P863_3_T", "D16")

d3 <- d2 %>% 
   filter(as_factor(LANGUAGE) == "English")

d3$P862_1_T <- as.character(d3$P862_1_T)
d3$P862_2_T <- as.character(d3$P862_2_T)
d3$P862_3_T <- as.character(d3$P862_3_T)
d3$P863_1_T <- as.character(d3$P863_1_T)
d3$P863_2_T <- as.character(d3$P863_2_T)
d3$P863_3_T <- as.character(d3$P863_3_T)
d3$D16 <- as.numeric(d3$D16)

d3 <- d3 %>%
   mutate(ideo_x = case_when(D16 <= 2 ~ "Cons",
                             D16 == 3 ~ "Mod",
                             D16 <= 5 ~ "Lib",
                             TRUE ~ NA_character_))

glimpse(d3)
```

Now we are ready to create our first word cloud.  Here we do just a single cloud.  First, we have to load the `tidytext` library which allows us to extract easily extract words from the open ended text using `unnest_tokens`, and then we also load the `wordcloud` library for obvious reasons.  

```{r}
library(tidytext)
library(wordcloud)
```

The first thing we'll need to do is filter out stop words, including custom ones. Stop words are words that are commonly used that we are NOT interested in analyzing.  The tidytext package comes with a set of preloaded stop words such as `c("the", "and", "or")` and so on.  Here's what that built in list looks like:

```{r}
stop_words
```

We can create our own list of custom stop words that we find in the data and do not want analyzed.  For this question, we probably want to filter out the word "president" because the question is specifically asking about how they would feel about her being president.  We also filter out "ummm" and "answer".  We turn that vector into a data frame that looks more like the pre-built stop words list, so that it will play nicely with it.

```{r}
custom_stops <- c("ummm", "answer", "president")

custom_stop_words <- tibble(word = custom_stops) %>%
   mutate(lexicon = "custom")

custom_stop_words
```

From there we are ready to begin manipulating the actual data to get it ready for our word cloud.  First, we filter out any NAs for ideology.  Second, respondents were able to list three separate answers and they are available in three different variables, so we combine all three responses into one master open-ended response that we can do a full analysis on.  

Next we clean up some of the typos in the data and also consolidate similar responses.  For example, we fix a misspelling of Obama's first name, and then recode variations of the word "woman" into the same word so that all of these items are rolled up into one word in the cloud.  This process is necessarily interative.  You'll run the word cloud, look at the results, see a misspelling and then add it to the statement, rerun the cloud, see two related words that need to be rolled up together, rerun the word cloud and so on, until you are satisfied.  

```{r}
# Suppose Hillary Clinton is elected president in 2016. In your view, what would be the best or most positive thing about a Hillary Clinton presidency? 
# Single word cloud
d4 <- d3 %>% 
   filter(is.na(ideo_x) == F) %>%
   mutate(clinton_oe = str_c(P862_1_T, ";", P862_2_T, ";", P862_3_T)) %>% 
   mutate(clinton_oe = str_replace_all(clinton_oe %>% tolower, "barrack", "barack"),
          clinton_oe = str_replace_all(clinton_oe %>% tolower, "women", "woman"),
          clinton_oe = str_replace_all(clinton_oe %>% tolower, "women's", "woman"),
          clinton_oe = str_replace_all(clinton_oe %>% tolower, "woman's", "woman"))

d4
```

Next we finally create the actual wordcloud.  To do so, we `unnest_tokens` within the variable `clinton_oe` and into a new variable called `word`.  We `anti_join` on the `stop_words` library AND the `custom_stop_words` library.  We then call a `count` function to sum up the number of times each word occurs, but use our survey weight variable for that counting, and sort accordingly.

Then lastly, we call the wordcloud function.  This has a number of parameters that we can customize, the maximum number of words to display, the size of those words, whether they are in a random order, whether they are rotated and finally, what color they should be.  

```{r}
d4 %>% 
   unnest_tokens(word, clinton_oe) %>%
   anti_join(stop_words) %>%
   anti_join(custom_stop_words) %>%
   count(word, wt = WTFCTR, sort = T) %>%
   with(wordcloud(word, n, 
                  max.words = 50, 
                  scale=c(3, 0.6), 
                  random.order = F,
                  rot.per = 0,
                  color = "dark green"))
```

Next we'll do a comparison wordcloud which allows us to create a faceted wordcloud based on some categorical characteristic.  In this case, we'll do it on Ideology.  We have the same prep work, but this time the `count` function includes the variable we want to do a comparison on - `ideo_x`.  From there we need to do a few things to get the data into the right shape for a comparison cloud.  First, the comparison cloud requires that the compared items be in the columns, so we have to `pivot_wider` with the column names coming from `ideo_x` and the values coming from our `n` variable.  Also, the comparison cloud doesn't know what to do with NA values so we set `values_fill` to 0, replacing any NAs with zero.  Finally, the comparison cloud requires that the word data be the row name (so that it looks more like a term document matrix, which is what it is actually expecting), so we have to use the `column_to_rownames` function and set the rownames equal to `word`.  Finally, we can call `comparison.cloud` which has similar parameters to the previous one: we can set the colors, the scale, the maximum number of words, whether we want them in a random order, the title size and whether we want the words rotated.  

```{r}
# Suppose Hillary Clinton is elected president in 2016. In your view, what would be the best or most positive thing about a Hillary Clinton presidency? 
d4 %>% 
   unnest_tokens(word, clinton_oe) %>%
   anti_join(stop_words) %>%
   anti_join(custom_stop_words) %>%
   count(ideo_x, word, wt = WTFCTR, sort = T) %>%
   pivot_wider(names_from = ideo_x, values_from = n, values_fill = 0) %>% 
   column_to_rownames(var = "word") %>% 
   comparison.cloud(colors = c("red", "blue", "grey"),
                    scale = c(0.6,3),
                    max.words = 30,
                    random.order = F,
                    title.size = .5,
                    rot.per = 0)
```

Here we do the same things, but this time for the other open end question, "What would be the worst or most negative thing about a Hillary Clinton presidency?"

```{r}
# And, what would be the worst or most negative thing about a Hillary Clinton presidency?
# Single word cloud
d3 %>% 
   filter(is.na(ideo_x) == F) %>%
   mutate(clinton_oe_neg = str_c(P863_1_T, ";", P863_2_T, ";", P863_3_T)) %>% 
   mutate(clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "barrack", "barack"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "women", "woman"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "women's", "woman"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "woman's", "woman")) %>% 
   unnest_tokens(word, clinton_oe_neg) %>%
   anti_join(stop_words) %>%
   anti_join(custom_stop_words) %>%
   count(word, wt = WTFCTR, sort = T) %>%
   with(wordcloud(word, n, 
                  max.words = 50, 
                  scale=c(3, 0.6), 
                  random.order = F,
                  rot.per = 0,
                  color = "black"))
```

Same question, but now a comparison cloud:

```{r}
d3 %>% 
   filter(is.na(ideo_x) == F) %>%
   mutate(clinton_oe_neg = str_c(P863_1_T, ";", P863_2_T, ";", P863_3_T)) %>% 
   mutate(clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "barrack", "barack"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "women", "woman"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "women's", "woman"),
          clinton_oe_neg = str_replace_all(clinton_oe_neg %>% tolower, "woman's", "woman")) %>% 
   unnest_tokens(word, clinton_oe_neg) %>%
   anti_join(stop_words) %>%
   anti_join(custom_stop_words) %>%
   count(ideo_x, word, wt = WTFCTR, sort = T) %>%
   pivot_wider(names_from = ideo_x, values_from = n, values_fill = 0) %>% 
   column_to_rownames(var = "word") %>% 
   comparison.cloud(colors = c("red", "blue", "grey"),
                    scale = c(0.6,3),
                    max.words = 30,
                    random.order = F,
                    title.size = .5,
                    rot.per = 0)
```

### Network Analysis

For our purposes we will be using the `ggraph` package which allows for the visualization of network graphs using methods familiar to `ggplot2`.  This package builds on top of `tidygraph` which is in turn built off of the `igraph` package.  `igraph` is the package which allows you to calculate statistics on network datasets, while `tidygraph` provides tidy wrappers to it.  

```{r}
library(igraph)
library(tidygraph)
library(ggraph)
```

There is an important distinction between traditional ggplot plots and ggraph plots: network data is fundamentally different from tabular data, and most network visualizations don’t concern themselves with mapping variables to x and y aesthetics since they are concerned with showing the network topology more than relations between two variables. In order to show network topology the concept of layouts are employed. Layouts are algorithms that use the network structure to calculate (often arbitrary) x and y values for each node that can then be used for visualization purposes. To put it in another way, when plotting tabular data the x and y aesthetics are almost always mapped to existing variables in the data (or statistical transformations of existing data) whereas when plotting network data x and y are mapped to values derived from the topology of the network and which are by themselves meaningless.

In other words, in general, network graphs do not have coordinates. Thus, there is no right way to draw a network graph. Visualizing a graph is more art than science, but several graph layout algorithms are popular.

For this demo, we'll be working with node and edge data from the movie Star Wars: A New Hope (Episode 4).  The data is built by processing the script for the movie and counting the number of shared scenes between characters.  

```{r}
# big thanks to https://dgarcia-eu.github.io/SocialDataScience/5_SocialNetworkPhenomena/057_Tidygraph2/tidygraph2.html
sw_edges <- read_csv("sw_edges.csv")
sw_nodes <- read_csv("sw_nodes.csv")
```

Notice what these data sets look like.  First, here are the nodes, or the characters from the movie, which have an id associated with them, and they have a categorical designation as to whether they are "good guys" or "bad guys":

```{r}
glimpse(sw_nodes)
```

Now, here are the edges, in which there is a source and a target (two characters sharing a scene), a weight that in this instance is a count of the number of shared scenes, and then a categorical variable of the interaction between these two characters - that is, whether they are engaging as allies or adversaries.   

```{r}
glimpse(sw_edges)
```

In order to visualize these as a network map, the first thing you need to do is convert the two separate data frames into an `tidygraph` object.  For this, we can use the `tbl_graph` function and pass it objects for our nodes, our edges, and tell it that the network is undirected.  You'll see that the new object is of type = `tbl_graph` and that it contains 2 components, node data and edge data.  All the original data is preserved.

```{r}
swnet <- tbl_graph(sw_nodes, sw_edges, directed = FALSE)

swnet
```

There are two ways you can generate a quick plot of the data just to see what you are dealing with.  There is the base R function of `plot` and also a special ggraph function `autograph` that serves the same function:

```{r}
# quickplot of network map
plot(swnet)
```

```{r}
# quickplot with ggraph function, meant for quick overviews
autograph(swnet)
```

You initialize the plot with a `ggraph()` call. The first argument is the data, which can be a tbl_graph or any object convertible to one.  The second argument is the layout argument.  If you do not specify the format, it defaults to a `stress` layout.  From there, you'll note that the syntax is very similar to ggplot, but with new types of geoms, each related to edges and nodes. You can, for example, draw a plot with just nodes:

```{r}
# draw basic network map
ggraph(swnet) + 
   geom_node_point() +
   theme_graph()
```

Or you can draw it with just edges:

```{r}
# draw basic network map
ggraph(swnet) + 
   geom_edge_link() + 
   theme_graph()
```

But most frequently, you'll be drawing it with edges AND nodes.  Below, I've done exactly that AND I've specified a different kind of layout, a forced layout:

```{r}
# draw basic network map
ggraph(swnet, "fr") + 
   geom_edge_link() + 
   geom_node_point() +
   theme_graph()
```

This isn't all that helpful, as we don't know which nodes are which.  Here's where you can start interacting with it using the tools you've learned as part of ggplot.  For example, we can add a new type of text geom that is specific to nodes, and we have to map its label aesthetic to `name`:

```{r}
ggraph(swnet, "fr") + 
   geom_edge_link() + 
   geom_node_point() + 
   geom_node_text(aes(label = name), repel=TRUE) +
   theme_graph()
```

That is much better!  But we can also add other aesthetics, like mapping color to whether or not they are good guys:

```{r}
ggraph(swnet, "fr") + 
   geom_edge_link() + 
   geom_node_point(aes(color = good)) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("black", "green")) +
   theme_graph()
```

As I said before, there are a number of different layouts we can use. Here are a few of them:

```{r}
# vary layout methods
# graphopt, kk, grid, circle, tree, unrooted
ggraph(swnet, "graphopt") +
   geom_edge_link() +
   geom_node_point(aes(color = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

```{r}
ggraph(swnet, "kk") +
   geom_edge_link() +
   geom_node_point(aes(color = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

```{r}
ggraph(swnet, "grid") +
   geom_edge_link() +
   geom_node_point(aes(color = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("green", "black")) +
   theme_graph()

```

```{r}
ggraph(swnet, "tree") +
   geom_edge_link() +
   geom_node_point(aes(color = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

```{r}
ggraph(swnet, "unrooted") +
   geom_edge_link() +
   geom_node_point(aes(color = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

Using the tidygraph library, you can calculate a number of metrics, which can in turn be used in your visualization.  Below, we identify natural "communities" in the data, using a community detection algorithmn to find densely connected subgraphs.  Here we use the Louvain algorithm, which is a fast and common way to find a division into communities that maximizes modularity.  This uses the `group_louvain()` function, and we convert the output to a character so that it can display properly as a categorical variable.  

Notice here the new verb that we are using, `activate`.  This is specific to tidygraph and is used to declare whether you are going to be conducting an operation on the nodes of the graph object or the edges of the graph object.  Since we are calculating a community, it is being done on our nodes.  

We can view the results of this calculation by activating the nodes and then using `as_tibble`:

```{r}
swnet <- swnet %>% 
   activate(nodes) %>% 
   mutate(community = as.character(group_louvain()))

# review it
swnet %>% 
   activate(nodes) %>% 
   as_tibble()
```

And now that we have the community designation, we can plot it:

```{r}
# plot it
ggraph(swnet, "fr") +
   geom_edge_link() +
   geom_node_point(aes(color = community), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   theme_graph()
```

We can continue to improve the graph by varying the encodings.  Here we keep the community coloring, but add a shape aesthetic that is mapped to what side they are on:

```{r}
ggraph(swnet, "fr") +
   geom_edge_link() +
   geom_node_point(aes(color = community, shape = good), size = 3) + 
   geom_node_text(aes(label = name), repel=TRUE) +
   theme_graph()
```

We can also use the weight information (the number of shared scenes) in the visualization.  Below, we attach weight to the width aesthetic, but also set the alpha to a very low value, since it will undoubtedly get messy in the middle of that graph.

```{r}
# vary the width size by the weight variable, with alpha set appropriately
ggraph(swnet, "fr") +
   geom_edge_link(aes(width = weight), alpha = 0.25) +
   geom_node_point(aes(color = good)) +
   geom_node_text(aes(label = name), repel = TRUE) +
   theme_graph()
```

This is okay, but maybe we can try something else that will work better, like setting the alpha as an aesthetic and mapping it to weight:

```{r}
ggraph(swnet, "fr") +
   geom_edge_link(aes(alpha = weight)) +
   geom_node_point(aes(color = good)) +
   geom_node_text(aes(label = name), repel = TRUE) +
   theme_graph()
```

You can also use faceting in ggraph, but the function calls are specialized to whether you are faceting by a node element, a edge element, or both.  Here we are faceting by the node attribute of `good`:

```{r}
# facet by node characteristic
ggraph(swnet, "fr") +
   geom_edge_link() +
   geom_node_point(aes(color = good)) +
   geom_node_text(aes(label = name), repel = TRUE) +
   facet_nodes(~good)
```

Here we facet by the edge characteristic of whether the interaction was between allies or adversaries:

```{r}
ggraph(swnet, "fr") +
   geom_edge_link() +
   geom_node_point(aes(color = good)) +
   geom_node_text(aes(label = name), repel = TRUE) +
   facet_edges(~exchange)
```

And finally you can facet by both a node and edge characteristic:

```{r}
ggraph(swnet, "fr") +
   geom_edge_link() +
   geom_node_point(aes(color = good)) +
   geom_node_text(aes(label = name), repel = TRUE) +
   facet_graph(exchange~community)
```

Here's another example of making a tidygraph calculation and then plotting it, this time using a measure of centrality:

```{r}
# calculate centrality
swnet <- swnet %>% 
   activate(nodes) %>% 
   mutate(centrality = centrality_betweenness())

# display it
swnet %>% 
   activate(nodes) %>% 
   arrange(desc(centrality)) %>% 
   as_tibble()
```

Now we can plot it, mapping the size aesthetic to that centrality number:

```{r}
ggraph(swnet, layout = "fr") + 
   geom_edge_link() + 
   geom_node_point(aes(size = centrality)) + 
   geom_node_text(aes(label = name), repel=TRUE)
```

There are some other layouts that are more distinctive, like a linear one:

```{r}
# linear 
ggraph(swnet, "linear") +
   geom_edge_arc(aes(color = exchange)) +
   geom_node_text(aes(label = name), repel=TRUE) +
   geom_node_point(aes(color = good), size = 3) + 
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

We can also take this and turn it into a chord diagram, by setting the circular parameter to TRUE on a linear layout:

```{r}
# chord diagram
ggraph(swnet, "linear", circular = TRUE) +
   geom_edge_arc(aes(color = exchange)) +
   geom_node_text(aes(label = name), repel=TRUE) +
   geom_node_point(aes(color = good), size = 3) + 
   scale_color_manual(values = c("green", "black")) +
   theme_graph()
```

